Updated Criteria for the Diagnostic Procedure for Parkinson’s Disease Dementia on Level I and their Validity in Deep Brain Stimulation Cohort

### 2025-05-12 _manuscript/manuscript.qmd
Last few changes in the text, now up to JM to check the text for  possible baloney :)

Parkinson’s disease (PD) is a neurodegenerative disorder characterised by a progressive onset of motor symptoms, including rigidity, bradykinesia, postural instability and resting tremor. Beyond motor symptoms, patients routinely suffer from non-motor impairments [@postuma2015], such as cognitive decline. This decline might peak in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

Despite its clinical relevance, the diagnosis of PDD remains complex. The original International Parkinson and Movement Disorder Society (MDS) diagnostic criteria [@dubois2007], first formalized in 2007, were heavily influenced by frameworks established primarily for Alzheimer’s disease (AD). Although these criteria provided a valuable initial foundation, they lacked the specificity required to capture the distinct pathophysiological and cognitive features of PD-related dementia [@emre2007; @yamashita2023].

To enhance diagnostic accuracy, the MDS introduced a two-level system for Parkinson’s disease dementia. Level I criteria are designed for clinical use, relying on brief cognitive assessments and clinical judgment. Level II criteria involve comprehensive neuropsychological testing across multiple domains, providing greater diagnostic detail for research or specialized settings [@emre2007].

One notable feature of the original Level I criteria was the provision of an algorithm that allowed for flexibility in test selection [@dubois2007]. Specifically, clinicians could choose between months reversed or seven backwards for attention assessment, lexical fluency or clock drawing for executive function evaluation. Moreover, the critaria included MMSE pentagons for evaluation of visuospatial ability, and three-word recall for memory assessment. Agreement across different criteria allows for the parallel computation of inter-rater reliability , which, in turn, facilitates the calculation of construct validity [@conway1995] further strengthening the diagnostic framework for PDD.

According to a recent metanalysis, the epidemiological estimates of PDD prevalence among individuals with PD vary widely, ranging from 14% to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients\` sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline [@oh2016; @rana2012].

Currently, efforts are focused on refining the PDD diagnostic framework to improve its consistency and applicability in both research and clinical contexts [@kulisevsky2024].

The present study evaluates the diagnostic concordance between the original Level I PDD criteria, as established by the MDS Task Force [@emre2007; @dubois2007] and criteria inspired by the recent call for change [@kulisevsky2024] within a cohort of PD patients considered for DBS. Furthermore, both sets of criteria are compared to PDD diagnosed on Level II. The study aims to address following research objectives (RO): (RO1) To estimate the Parkinson’s disease dementia (PDD) rates and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components that contribute to variability in PDD classification across the applied criteria.

@@

This study comprehensively examined the diagnostic variability and validity of Level I criteria for PDD in DBS treatment candidates. By systematically comparing 68 diagnostic algorithms based on MMSE, MoCA, sMoCA, and Level II criteria, our findings highlight the significant influence of algorithm design.

@@

Our results showed a wide range in estimated PDD prevalence across 68 algorithms, ranging from 2.00% to 16.75%. Notably, estimates were substantially lower when using FAQ item 9 alone versus the full FAQ scale. This discrepancy underscores the diagnostic weight of IADL assessment. Our PDD rates were consistently lower compared to previous studied, demonstrating wide variability in reported PDD rate among PD patients. For instance, retrospective study reported a PDD rate of 19.7% [@rana2012], while a clinical investigation found even higher rate of around 30% [@aarsland2011]. A recent complex meta-analysis synthesizing global data placed the pooled rate at 26.30% [@sousa2022]. **The discrepancy is likely attributable to our DBS specific cohort, which differs in important characteristics, such as younger age of the patients [@rana2012]. / Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in sample characteristics, diagnostic criteria, and methodology. [^3]**

@@

Our pairwise comparison of diagnostic algorithms showed that agreement was notably stronger among those using the same IADL operationalization (κ range from 0.75, SD = 0.13 to 0.86, SD = 0.09) compared to those using different IADL definitions (κ = 0.34, SD = 0.08). This demonstrates that even subtle methodological differences can yield substantially divergent diagnostic outcomes. Such findings are critical for clinicians relying on Level I criteria for eligibility decisions, as the choice of algorithm could lead to contradictory classifications of PDD.

## Predictive Validity Comaprison With Level II Criteria

When using Level II diagnosis as the reference, MoCA-based Level I algorithms, particularly those using Clock Drawing used to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernize PDD diagnostic frameworks [@kulisevsky2024], favoring MoCA-derived components and broader functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing the nuanced cognitive deficits characteristic of PDD.

The advantage of MoCA-based algorithm was particularly evident when using the FAQ total score to define functional impairment in comparison with using FAQ item 9 only. In these cases, the best-performing algorithms achieved accuracy rates of 94–99%, together with balanced sensitivity and specificity, reinforcing the clinical utility.

@@

These findings have immediate implications for DBS candidacy assessment. Given that a diagnosis of PDD typically precludes surgical intervention, inconsistent application of diagnostic criteria can lead to inequitable access to the treatment. Our data emphasize the need for harmonized assessment protocols that balance sensitivity with specificity, particularly in high-stakes clinical decision-making.

Moreover, the relative success of multidomain assessments, such as Level II protocols or regression-based normative approaches, highlights the importance of psychometric rigor in diagnosing PDD. These approaches may serve as confirmatory tools in cases where Level I assessments yield ambiguous results.

@@

This study’s generalizability is limited by its retrospective design and the homogeneity of the DBS candidate cohort, which may not reflect broader PD populations with varying cognitive profiles. Additionally, reliance on Czech normative data may restrict international applicability, although it represents real-world clinical standards within the region [@simons2017].

@@

Future research should focus on prospective validation of updated Level I criteria against longitudinal functional outcomes and neurobiological markers. Additionally, the development of adaptive diagnostic algorithms that can integrate performance across cognitive, functional, and psychiatric domains may enhance diagnostic sensitivity while reducing the risk of over-classification.

@@

This study systematically investigated the application of multiple Level I diagnostic criteria for PDD within a cohort of patients considered for DBS. The findings reveal substantial variability in PDD rates estimates, strongly influenced by the choice of cognitive screening instruments and the operationalization of functional impairment. The divergence observed across algorithms demonstrates the sensitivity of diagnostic outcomes to seemingly minor methodological choices.

The proposed revisions to the diagnostic framework [@kulisevsky2024] criteria offer enhanced sensitivity by leveraging MoCA-based components and broader IADL assessments. Conversely, overly conservative criteria, such as reliance on pill questionnaire (i.e. FAQ item 9 equivalence) may fail to detect meaningful functional decline and thus under-identify true cases of PDD presence.

Ultimately, this study contributes to the improving landscape of PDD diagnostics by offering empirical evidence for the refinement of Level I criteria and reinforcing the value of psychometric rigor in clinical neuropsychology. Future research should validate these operationalizations longitudinally and explore integration with biomarkers to further enhance diagnostic reliability.

>>>>

Parkinson’s disease (PD) is a neurodegenerative disorder defined by a gradual onset of motor impairment, including rigidity, bradykinesia, postural instability and resting tremor. Additionally, patients frequently suffer from a range of non-motor impairments [@postuma2015], particularly cognitive decline. This state might result in Parkinson’s disease dementia (PDD) in a subset of patients [@meireles2012].

Despite its clinical relevance, the diagnosis of PDD remains complex. The original criteria formalized in 2007 by the International Parkinson and Movement Disorder Society (MDS) [@dubois2007] were heavily influenced by frameworks established primarily for Alzheimer’s disease (AD). Although these criteria provided a valuable base, they lacked the specificity to detect the specific cognitive aspects of PDD [@emre2007; @yamashita2023].

In order to increase diagnostic accuracy, the MDS also developed a levelled system for PDD detection. Level I criteria are designed primarily for clinical use, consisting of brief cognitive assessments and relying on clinical judgement. In contrast, Level II criteria use comprehensive neuropsychological testing across cognitive domains to provide greater diagnostic specificity [@emre2007].

One specific feature of the original Level I criteria was the presence of an algorithm allowing flexibility in test selection. The criteria included MMSE pentagons for evaluation of visuospatial ability, three-word recall for memory assessment,  months reversed or seven backwards for attention assessment and lexical fluency or clock drawing for executive function assessment [@dubois2007].

According to a recent meta-nalysis, the PDD prevalence estimates vary widely, ranging from 14% up to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients` sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline [@oh2016; @rana2012]. Currently, efforts are focused on refining the PDD diagnostic framework to improve its consistency and applicability in both research and clinical contexts [@kulisevsky2024].

The present study evaluates the diagnostic concordance between the original Level I PDD criteria, as established by the MDS Task Force [@emre2007; @dubois2007] and criteria inspired by the recent call for change [@kulisevsky2024] within a cohort of PD patients considered for DBS. Furthermore, both sets of criteria are compared to PDD diagnosed on Level II. The study aims to address the following research objectives (RO): (RO1) To estimate the Parkinson’s disease dementia (PDD) rates and evaluate the diagnostic variability and concordance across different PDD criteria. (RO2) To identify specific diagnostic components contributing to PDD classification variability across the applied criteria.

@@

This study examined the diagnostic variability and validity of Level I criteria for PDD, specifically in candidates for DBS treatment. By systematically comparing 68 diagnostic algorithms based on MMSE, MoCA, sMoCA, and Level II criteria, our findings highlight the significance of algorithm design.

@@

Our results showed a wide range in estimated PDD prevalence across the total of 68 algorithms, ranging from 2.00% to 16.75%. Estimates reached lower rates when using solely FAQ item 9 in comparison with the full FAQ scale. This discrepancy highlights the diagnostic importance of IADL assessment. Our overall PDD rates were consistently lower compared to previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rates, reaching up to 30% [@aarsland2011]. A recent complex meta-analysis synthesizing global data placed the rate at 26.30% [@sousa2022]. **The discrepancy is likely attributable to our DBS-specific cohort, which differs in important characteristics, such as the younger age of the patients [@rana2012]. / Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in sample characteristics, diagnostic criteria, and methodology. [^3]**

@@

Our pairwise comparison of diagnostic algorithms showed that agreement was notably stronger among those using the same IADL operationalization (κ range from 0.75, SD = 0.13 to 0.86, SD = 0.09) compared to those using different IADL definitions (κ = 0.34, SD = 0.08). This demonstrates that even slight methodological differences can yield divergent diagnostic outcomes. Such findings are critical for clinicians relying on Level I criteria for eligibility decisions, as the choice of algorithm could lead to contradictory classifications of PDD.

## Predictive Validity Comparison With Level II Criteria

When using Level II diagnosis as the reference, MoCA-based Level I algorithms, particularly those using Clock Drawing to assess executive function, demonstrated the highest predictive accuracy. This supports recent proposals to modernize PDD diagnostic frameworks [@kulisevsky2024], favoring MoCA-derived components and broader functional assessment tools. In contrast, MMSE-based algorithms consistently underperformed, suggesting limited sensitivity in capturing the nuanced cognitive deficits characteristic of PDD.

The advantage of MoCA-based algorithm was evident when using the FAQ total score to define functional impairment in comparison with using solely FAQ item 9. In these cases, the best-performing algorithms achieved accuracy rates of 94–99%, together with high sensitivity and specificity.

@@

Our findings have immediate implications for DBS candidacy assessment. Given that a diagnosis of PDD typically excludes surgical intervention, inconsistent application of diagnostic criteria can lead to unbalanced access to the treatment.

Moreover, the relative success of multidomain assessments, such as Level II protocols, highlights the importance of psychometric precision. This approach may serve as a confirmatory tool in cases where Level I assessments provide ambiguous results.

@@

This study’s generalizability is limited by its retrospective design and the homogeneity of the DBS candidate cohort, which may not reflect broader PD populations with varying cognitive profiles. Moreover, use of the Czech normative data may limit international applicability, although our results represent clinical standards in the region [@simons2017].

@@

Future research should focus on the validation of updated Level I criteria using longitudinal data and neurobiological markers. Furthermore, the development of adaptive diagnostic algorithms that can combine performance across cognitive and functional domains may enhance diagnostic sensitivity while reducing the risk of over-classification of PDD.

@@

This study systematically investigated the application of multiple Level I diagnostic criteria for PDD within a cohort of patients considered for DBS tretment. Our results show variability in PDD rate estimates, strongly influenced by the choice of cognitive screening instruments and the operationalization of functional impairment. The divergence observed across algorithms demonstrates the sensitivity of diagnostic outcomes to seemingly negligible methodological choices.

The proposed changes of the current diagnostic criteria [@kulisevsky2024] offer heightened sensitivity by proposing the use of MoCA-based components and broader IADL assessments. Overly conservative criteria, such as reliance on pill questionnaire (i.e. FAQ item 9 equivalence), may fail to detect meaningful functional decline and thus under-identify true cases of PDD.

Conclusively, this study contributes to the improvement of PDD diagnostics by offering evidence for the clarification of Level I criteria and emphasizing the value of psychometric precision in clinical neuropsychology. Future research should validate the algorithms longitudinally and explore possible integration of biomarkers to further improve diagnostic reliability.
